Quantitative processing of broadband data as implemented in a scientific splitbeam echosounder

### 2022-04-25 Paper/Manuscript.tex
Edits for readability in Introduction

Active acoustics is an efficient method for remote sensing of marine ecosystems. These methods have been used to cover a wide range of spatial and temporal scales, ranging from tens of km \citep{makris_fish_2006} to small scale behaviour patterns \citep{klevjer_split-beam_2003}, and can provide information on key forage species, including fish and zooplankton, linking primary producers and top predators as well as other taxonomic groups important for the ecosystem functioning \citep{benoit-bird_ecological_2016}. As early as 1935, Oscars Sund observed the distribution of spawning cod in the Lofoten area using a single beam echosounder \citep{sund_echo_1935}. The method was further developed to map the abundance of fish, driven by the need for fisheries management \citep{Simmonds2005Fisheries}. More recently, the methods have been deployed on a wide range of platforms, including observatories, autonomous underwater vehicles \citep{fernandes_autonomous_2003}, uncrewed surface vehicles \citep{de_robertis_uncrewed_2021} and vessels of opportunity, observing a wide range of ecosystem processes across different spatial and temporal scales \citep{godo_marine_2014}.

Several different configurations of acoustic instruments are available, including multibeam sonars \citep{gerlotto_two_1999}, synthetic aperture sonar, omnidirectional sonars \citep{misund_improved_1996}, acoustic imaging sonars \citep{jaffe_ftv_1995} as well as single- and multibeam echosounders \citep{trenkel_new_2008}. Narrowband single beam echosounders have been used extensively in fisheries management and ecosystem monitoring, and methodology for using these systems are well developed. The ability for accurate calibration of the system has made these especially useful for providing data for fisheries management \citep{Simmonds2005Fisheries}.

Recent commercially available single beam echosounders produce pulses with a wide and continuous frequency range (broadband pulses), as opposed to conventional single frequency systems. This provides significantly better along-beam resolution, a higher signal to noise ratio than narrowband pulses \citep{Chu1998Application, ehrenbergFMSlideChirp2000}, and  improved frequency resolution for backscatter categorization \citep{korneliussen2018}.

Broad banded echosounders have been pursued actively for ecology in particularly based on the promise of improved categorization of species or group of species. Using high frequency broadband acoustics (150-600 kHz) Lavery et al \citep{lavery_measurements_2010} showed that broadband helped reduce the ambiguities in interpretation of acoustic scattering from zooplankton and oceanic microstructure. \citep{Stanton2012Resonance} used low frequency broadband pulses (1-6 kHz) with frequency content covering the resonance of swimbladdered fish, demonstrating both the spectral resolution and ability to classify size classes of fish within fish assemblages. \citep{blanluet_characterization_2019} used broadband, with ground truthing, to identify that the composition of two Sound Scattering Layers (SSL) detected in the Bay of Biscay in springtime was characterized and identifying fine scale heterogeneity within the SSLs. While unsuccessful in discriminating between several species of fishes with large  swimbladders during the Alaska pollock survey, \citep{bassett_broadband_2018} showed that broadband may be helpful in characterizing smaller fishes with swimbladder and euphausiids. \citep{benoit-bird_exploring_2020} was able to effectively discriminate three monospecific aggregations of species (hake, anchovy and krill) using broadband data (45-170 kHz), while attempts to classify using narrowband multifrequency failed.

The increased spatial resolution is another aspect of broadband acoustics that are actively being used both for improved categorization and increased ability to resolve targets near each other or near boundaries such as the seafloor. \citep{lavery2017} explored different broadband pulse shapes to increase successful resolution of single targets in proximity as well as near boundaries in tank experiments. By applying high frequency broadband pulses to fish-like artificial targets, \citep{kubilius_remote_2020} demonstrated the potential for acoustic sizing of individually resolved fishes in a controlled ex situ environment. Utilizing the increased spatial resolution \citep{hasegawa_situ_2021} were able to isolate single fishes and discriminate successfully between average frequency responses of walleye pollock and pointhead flounder.

An equally interesting outcome from the increased resolution and broad frequency bands is the ability to infer behavioural information on both single individuals, groups of individuals, and interaction between individuals or groups. \citep{Traykovski1998Effect} was able to extract orientation information about krill from using a combination of high frequency broadband acoustics and acoustic modelling. Fine-scale observations of single targets facilitated the study of interactions between prey and predators, capelin and cod, in the Barents sea (\citep{skaret_diel_2020}).

There have been several scientific broadband echosounder systems developed for laboratory use \citep{Conti2003Wide-bandwidth, Forland2014Scattering, chu1992}, some prototype or custom-made systems \citep{Zakharia1989Wide-band, Zakharia1996Wideband, Simmonds1996Species, Foote2005Measuring, Imaizumi2009Detection, Briseno-Avena2015ZOOPS, Barr2002Target} and some commercially available systems \citep{Gordon1998FishMASS, Zedel2003Acoustic, Stanton2010New, ehrenbergFMSlideChirp2000, dennyBroadbandAcousticFish1998}.

The theoretical foundation for signal processing comes from radar applications and is earlier documented for broadband echosounders \citep{stanton2008}. To tap the full potential of broadband echosounder data, further data processing methods are being developed \citep{lavery2017,bassett_broadband_2018}. Examples include adapting pulses to improve target separation as well as further parameters from the individual targets. Different send pulses can be envisioned as well as various methods for acoustic target classification. When translating equations into code, several choices must be made. Some of these are well founded in signal processing literature whereas others are of a more practical and ad-hoc nature. The latter is typically missing in conventional papers, and this makes it difficult to test the implementation of the signal processing methods in new echo sounders and post processing software.

The objective of this paper and associated code is to present a systematic and comprehensive description of the data processing steps for calibrated broadband echosounder data. The steps include pulse compression, target strength as a function of frequency [TS(f)], and volume backscattering strength as a function of frequency [Sv(f)]. The intention is that the code will be used to validate implementations in various relevant data processing software, provide a starting point for further developing signal processing, and serve as a learning resource for broadbanded acoustic data processing. We also note that the processing equations and methodology presented in this paper are similar to version 1.12.4 and earlier of the Simrad EK80 software.

@@

A set of equations for calculating calibrated, frequency-dependent, target strength and volume backscatter from broadband echosounder signals have been presented. Providing an open-source resource for anyone interested in learning and developing the techniques further.

>>>>

Active acoustics is an efficient method for remote sensing of marine ecosystems and is used to cover a wide range of spatial and temporal scales, ranging from tens of km \citep{makris_fish_2006} to small scale behaviour patterns \citep{klevjer_split-beam_2003}, and can provide information on key forage species, including fish and zooplankton, linking primary producers and top predators as well as other taxonomic groups important for the ecosystem functioning \citep{benoit-bird_ecological_2016}. As early as 1935, Oscar Sund observed the distribution of spawning cod in the Lofoten area using a single beam echosounder \citep{sund_echo_1935}. The method was further developed to map the abundance of fish, driven by the need for fisheries management \citep{Simmonds2005Fisheries}. More recently, active acoustics have been deployed on a wide range of platforms, including observatories, autonomous underwater vehicles \citep{fernandes_autonomous_2003}, uncrewed surface vehicles \citep{de_robertis_uncrewed_2021} and vessels of opportunity, observing a wide range of ecosystem processes across different spatial and temporal scales \citep{godo_marine_2014}.

Several different configurations of acoustic instruments are available, including multibeam sonars \citep{gerlotto_two_1999}, synthetic aperture sonar, omnidirectional sonars \citep{misund_improved_1996}, acoustic imaging sonars \citep{jaffe_ftv_1995} as well as single- and multibeam echosounders \citep{trenkel_new_2008}. Narrowband single beam echosounders have been used extensively in fisheries management and ecosystem monitoring, and the methodologies for using these systems are well developed. The ability for accurate calibration of the system has made these especially useful for providing data for fisheries management \citep{Simmonds2005Fisheries}.

Recent commercially available single beam echosounders produce pulses with a wide and continuous frequency range (broadband pulses), compared to conventional single frequency systems. This provides significantly better along-beam resolution, a higher signal to noise ratio than narrowband pulses \citep{Chu1998Application, ehrenbergFMSlideChirp2000}, and improved frequency resolution for backscatter categorization \citep{korneliussen2018}.

Broadband echosounders have been actively investigated for ecological research, based on the expectation of improved categorization of species or group of species. Using high frequency broadband acoustics (150-600 kHz) \citet{lavery_measurements_2010} showed that broadband signals helped to reduce the ambiguities in interpretation of acoustic scattering from zooplankton and oceanic microstructure. \citet{Stanton2012Resonance} used low frequency broadband pulses (1-6 kHz) that included the resonance frequency of swimbladdered fish, demonstrating both the spectral resolution and ability to classify size classes of fish within fish assemblages of broadband signals. \citet{blanluet_characterization_2019} used broadband, with ground truthing, to identify that the composition of two sound scattering layers (SSL) detected in the Bay of Biscay in springtime was able to be characterized, and also identified fine scale heterogeneity within the SSLs. While unsuccessful at discriminating between several species of fishes with large swimbladders during the Alaska pollock survey, \citet{bassett_broadband_2018} showed that broadband signals may be helpful in characterizing smaller fishes with swimbladders and euphausiids. \citet{benoit-bird_exploring_2020} was able to effectively discriminate three monospecific aggregations of species (hake, anchovy and krill) using broadband signals (45-170 kHz), while attempts to classify using multiple narrowband signals failed.

The increased spatial resolution is another aspect of broadband acoustics that is being actively used both for improved categorization and the increased ability to resolve targets near each other or near boundaries such as the seafloor. \citet{lavery2017} explored different broadband pulse shapes to increase the ability to resolve adjacent single targets as well as near boundaries in tank experiments. By applying high frequency broadband pulses to fish-like artificial targets, \citet{kubilius_remote_2020} demonstrated the potential for acoustic sizing of individually resolved fishes in a controlled ex situ environment. Utilizing the increased spatial resolution \citet{hasegawa_situ_2021} were able to isolate single fishes and discriminate successfully between average frequency responses of walleye pollock and pointhead flounder.

An equally interesting outcome from the increased resolution and broad frequency bands is the ability to infer behavioural information on both single individuals, groups of individuals, and interaction between individuals or groups. \citet{Traykovski1998Effect} was able to extract orientation information about krill by using a combination of high frequency broadband acoustics and acoustic modelling. Fine-scale observations of single targets facilitated the study of interactions between prey and predators (capelin and cod) in the Barents sea \citep{skaret_diel_2020}.

There have been several scientific broadband echosounder systems developed for laboratory use \citep{Conti2003Wide-bandwidth, Forland2014Scattering, chu1992}, some prototype or custom-made systems \citep{Zakharia1989Wide-band, Zakharia1996Wideband, Simmonds1996Species, Foote2005Measuring, Imaizumi2009Detection, Briseno-Avena2015ZOOPS, Barr2002Target} and some commercially available systems \citep{Gordon1998FishMASS, Zedel2003Acoustic, Stanton2010New, ehrenbergFMSlideChirp2000, dennyBroadbandAcousticFish1998}. Commercially available systems are now produced by several manufacturers.

The theoretical foundation for signal processing comes from radar applications and has been documented for broadband echosounders \citep{stanton2008}. To tap the full potential of broadband echosounder data, further data processing methods are being developed \citep{lavery2017,bassett_broadband_2018}. Examples include adapting pulses to improve target separation as well as further parameters from the individual targets. Different transmission pulse forms can be envisioned as well as various methods for acoustic target classification. When translating equations into computer processing code, several choices must be made - some of these are well founded in signal processing literature whereas others are of a more practical and ad-hoc nature. The latter is typically missing in the literature, making it difficult to test the implementation of the signal processing methods in new echo sounders and post processing software.

The objective of this paper and associated code is to present a systematic and comprehensive description of the data processing steps for calibrated broadband echosounder data. The steps include pulse compression, target strength as a function of frequency [$\ts(\freqsym)$], and volume backscattering strength as a function of frequency [$\sv(\freqsym)$]. The intention is that the code will be used to validate implementations in various relevant data processing software, provide a starting point for further developing active acoustic broadband signal processing, and to serve as a learning resource.

@@

A set of equations for calculating calibrated, frequency-dependent, target strength and volume backscatter from broadband echosounder signals have been presented. Providing an open-source resource for anyone interested in learning and developing the techniques further. We note that the processing equations and methodology presented in this paper are similar to version 1.12.4 and earlier of the Simrad EK80 software.
